---
title: "INFO 523 – Online Retail Data Mining Project"
author: "Rishabh Bhattad"
teammate: "Khushi Chavan"
output:
  pdf_document: default
  word_document: default
---

# Introduction

This project uses the Kaggle Online Retail dataset to perform association rule mining, clustering, and classification.

# Check and install the required packages only
```{r}
install_package <- c("tidyverse", "lubridate", "arules", "arulesViz", "cluster", "factoextra", "caret", "reshape2", "pROC", "randomForest", "ggplot2")

for(ip in install_package){
  if(!require(ip, character.only = TRUE)) 
    install.packages(ip)
}
```

```{r}

library(tidyverse)
library(lubridate)
library(arules)
library(arulesViz)
library(cluster)
library(factoextra)
library(caret)
library(reshape2)
library(pROC)
library(randomForest)
library(ggplot2)

set.seed(42)
```

# Load Data

```{r}
retail_raw <- read_csv("data/online_retail.csv")
glimpse(retail_raw)
```

# Basic Cleaning

```{r}
summary(retail_raw)
sum(is.na(retail_raw$CustomerID))
```

# Data Cleaning & Feature Engineering

```{r}
retail_clean <- retail_raw %>%
  # remove missing customer IDs
  filter(!is.na(CustomerID)) %>%
  #remove negative quantities and prices
  filter(Quantity > 0, UnitPrice > 0) %>%
  # remove cancelled invoices
  filter(!str_starts(InvoiceNo, "C")) %>%
  
  mutate(
    InvoiceDate = lubridate::parse_date_time(InvoiceDate,
                                             orders = c("d/m/Y H:M", "d/m/Y H:M:S",
                                                        "m/d/Y H:M", "m/d/Y H:M:S")),
    InvoiceYear = lubridate::year(InvoiceDate),
    InvoiceMonth = lubridate::floor_date(InvoiceDate, unit = "month"),
    TotalPrice = Quantity * UnitPrice
  )

# check cleaned data
glimpse(retail_clean)
summary(retail_clean$Quantity)
summary(retail_clean$UnitPrice)
sum(is.na(retail_clean$CustomerID))
n_distinct(retail_clean$CustomerID)
```

#Customer level aggregation per customer
```{r}
cust_level_aggregation <- retail_clean %>%
  group_by(CustomerID) %>%
  summarise(
    total_spent = sum(TotalPrice),
    total_qty = sum(Quantity),
    num_orders = n_distinct(InvoiceNo),
    avg_order = mean(TotalPrice),
    avg_items_per_order = total_qty/num_orders,
    recency = as.numeric(Sys.Date() - max(InvoiceDate)),
    .groups = "drop"
  )

cust_level_aggregation
```

#Exploratory Data analysis

```{r}
# Top countries by revenue
top_countries <- retail_clean %>%
  group_by(Country) %>%
  summarise(total_revenue = sum(TotalPrice)) %>%
  arrange(desc(total_revenue)) %>%
  head(10)

top_countries_plot <- ggplot(top_countries, aes(x = reorder(Country, total_revenue), y = total_revenue)) +
  geom_col(fill = "#1B9E77") +
  coord_flip() +
  labs(title = "Top 10 Countries", x = "Country", y = "Revenue") +
  theme(text = element_text(size = 25))

png("plots/top_countries_plot.png", width = 1200, height = 900)
print(top_countries_plot)
dev.off()
```

```{r}
# Top products by revenue
top_products <- retail_clean %>%
  group_by(Description) %>%
  summarise(total_revenue = sum(TotalPrice)) %>%
  arrange(desc(total_revenue)) %>%
  head(10)

top_products_plot <- ggplot(top_products, aes(x = reorder(Description, total_revenue), y = total_revenue)) +
  geom_col(fill = "#28B463") +
  coord_flip() +
  labs(title = "Top 10 Products", x = "Product", y = "Revenue") +
  theme(text = element_text(size = 25))

png("plots/top_products_plot.png", width = 1200, height = 900)
print(top_products_plot)
dev.off()
```

```{r}
# Monthly revenue pattern
monthly_revenue <- retail_clean %>%
  group_by(InvoiceMonth) %>%
  summarise(monthly_revenue = sum(TotalPrice))

Monthly_revenue_pattern <- ggplot(monthly_revenue, aes(x = InvoiceMonth, y = monthly_revenue)) +
  geom_line(color = "#8E44AD", size = 1.2) +
  geom_point(color = "#8E44AD", size = 2) +
scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(title = "Monthly revenue pattern",
       x = "Month", y = "Revenue") +
  theme(text = element_text(size = 25))

png("plots/Monthly_revenue_pattern.png", width = 1200, height = 900)
print(Monthly_revenue_pattern)
dev.off()
```

# Association Rule Mining
## 1. Apriori Algorithm

```{r}
# filter UK data
# We focus on UK customers because the dataset is heavily dominated by UK transactions.
# Using only UK ensures cleaner, denser association rules without noise from countries with very few orders.
retail_uk <- retail_clean %>%
  filter(Country == "United Kingdom")

# build invoice-item pairs
trans_df <- retail_uk %>%
  select(InvoiceNo, Description) %>%
  distinct()

# make transactions
trans_list <- split(trans_df$Description, trans_df$InvoiceNo)
retail_trans <- as(trans_list, "transactions")

# show transaction summary
retail_trans
summary(retail_trans)

# Run Apriori algorithm
rules <- apriori(
  retail_trans,
  parameter = list(supp = 0.001, conf = 0.5, minlen = 2)
)

# top rules
summary(rules)
rules_top <- sort(rules, by = "lift", decreasing = TRUE)
inspect(head(rules_top, 20))

# barplot of top rules
rules_df <- as(rules_top[1:15], "data.frame")

p <- ggplot(rules_df, aes(x = reorder(rules, lift), y = lift)) +
  geom_col(fill = "#2E86C1") +
  coord_flip() +
  labs(
    title = "Top 15 Association Rules by Lift",
    x = "Rule",
    y = "Lift"
  ) +
  theme_minimal(base_size = 16)

p

# save plot
png("plots/assoc_rules_barplot.png", width = 1400, height = 1000)
print(p)
dev.off()
```

## 2. ECLAT Algorithm

```{r}
rules_eclat <- eclat(
  retail_trans,
  parameter = list(supp = 0.001, minlen = 2)
)

inspect(head(rules_eclat, 15))

# convert to dataframe
rules_eclat_df <- as(rules_eclat[1:15], "data.frame")

# barplot for ECLAT
p_eclat <- ggplot(rules_eclat_df, aes(x = reorder(items, support), y = support)) +
  geom_col(fill = "#8E44AD") +
  coord_flip() +
  labs(
    title = "Top 15 ECLAT Rules by Support",
    x = "Itemset",
    y = "Support"
  ) +
  theme_minimal(base_size = 20)

p_eclat

png("plots/eclat_rules_barplot.png", width = 1400, height = 1000)
print(p_eclat)
dev.off()
```

# Customer Clustering

## 1. K-Means Clustering

```{r}

# remove extreme spenders (top 1%) for more stable clusters
q99 <- quantile(cust_level_aggregation$total_spent, 0.99, na.rm = TRUE)
cust_level_aggregation <- cust_level_aggregation %>%
  filter(total_spent <= q99)

# log transform to reduce skew
cust_feats <- cust_level_aggregation %>%
  mutate(
    log_total_spent = log1p(total_spent),
    log_total_qty = log1p(total_qty),
    log_num_orders = log1p(num_orders),
    log_avg_order = log1p(avg_order)
  ) %>%
  select(CustomerID, starts_with("log_"))

# scale data
cust_scaled <- scale(cust_feats[, -1])

# find 3 clusters
set.seed(42)
km <- kmeans(cust_scaled, centers = 3, nstart = 25)

# add cluster labels
cust_feats$cluster <- km$cluster

# assign readable segment names
cust_feats$segment <- dplyr::case_match(
  cust_feats$cluster,
  1 ~ "High-Value Frequent Buyers",
  2 ~ "Low-Value Casual Buyers",
  3 ~ "Large-Basket Buyers"
)

# view segment sizes
table(cust_feats$segment)

# barplot of cluster sizes
cluster_sizes <- as.data.frame(table(cust_feats$segment))
colnames(cluster_sizes) <- c("segment", "count")

p1 <- ggplot(cluster_sizes, aes(x = segment, y = count, fill = segment)) +
  geom_col() +
  labs(
    title = "Customer Segments (Defined by Purchasing Behavior)",
    x = "Segment",
    y = "Customer count",
    fill = "Segment"
  ) +
  theme_minimal(base_size = 12) +
  theme(text = element_text(size = 25))

p1

# save barplot
png("plots/cluster_sizes.png", width = 1200, height = 900)
print(p1)
dev.off()
```

## 2. Cluster Profiling

```{r}
# join cluster labels back to original (unlogged) features
cust_summary <- cust_level_aggregation %>%
  inner_join(cust_feats %>% select(CustomerID, segment), by = "CustomerID") %>%
  group_by(segment) %>%
  summarise(
    avg_total_spent = mean(total_spent),
    avg_total_qty = mean(total_qty),
    avg_num_orders = mean(num_orders),
    avg_order_value = mean(avg_order),
    .groups = "drop"
  )

# reshape for profile plot
cluster_long <- cust_summary %>%
  tidyr::pivot_longer(
    cols = c(avg_total_spent, avg_total_qty, avg_num_orders, avg_order_value),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = dplyr::recode(
      metric,
      "avg_total_spent" = "Total spent (GBP)",
      "avg_total_qty" = "Total quantity",
      "avg_num_orders" = "Number of orders",
      "avg_order_value" = "Avg order value (GBP)"
    )
  )

p2 <- ggplot(cluster_long, aes(x = segment, y = value, fill = segment)) +
  geom_col(position = "dodge") +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Customer Segment Profiles",
    x = "Segment",
    y = "Average metric value",
    fill = "Segment"
  ) +
  theme_minimal(base_size = 12)  +
  theme(
    text = element_text(size = 25),
    axis.text.x = element_blank(),      
    axis.ticks.x = element_blank()     
  )

p2

# save profile plot
png("plots/cluster_segment_profile.png", width = 1600, height = 900)
print(p2)
dev.off()
```

## 3. Hierarchical Clustering - Second Clustering Algorithm

```{r}
# use scaled log features for hierarchical clustering
d_matrix <- dist(cust_scaled, method = "euclidean")
# create full hierarchical model (needed for cutree)
hc <- hclust(d_matrix, method = "ward.D2")

# assign clusters from hierarchical model
hc_groups <- cutree(hc, k = 3)

# build cluster-wise averages for heatmap
heatmap_data <- as.data.frame(cust_scaled)
heatmap_data$cluster <- factor(hc_groups)

cluster_means <- heatmap_data %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean)) %>%
  rename_with(~ gsub("log_", "", .), -cluster)

# convert to long format for ggplot
heatmap_long <- melt(cluster_means, id.vars = "cluster")

# rename clusters to be more interpretable
heatmap_long$cluster <- dplyr::recode(
  heatmap_long$cluster,
  "1" = "Cluster 1: High Activity",
  "2" = "Cluster 2: Moderate Activity",
  "3" = "Cluster 3: Low Activity"
)

heatmap_long$variable <- dplyr::recode(
  heatmap_long$variable,
  "total_spent" = "Total Spent (GBP)",
  "total_qty" = "Total Quantity Purchased",
  "num_orders" = "Number of Orders",
  "avg_order" = "Average Order Value (GBP)"
)

# ggplot heatmap
p_hc <- ggplot(heatmap_long, aes(x = variable, y = cluster, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#EAF2F8", high = "#1B4F72") +
  labs(
    title = "Hierarchical Clustering – Average Feature Levels per Cluster",
    subtitle = "Clusters are labeled to indicate relative activity based on spending, quantity, and orders",
    x = "Customer Feature",
    y = "Hierarchical Cluster",
    fill = "Avg\nValue"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 12)
  ) +
  theme(text = element_text(size = 25))

p_hc

# save heatmap
png("plots/hierarchical_clustering_heatmap.png", width = 1400, height = 1000)
print(p_hc)
dev.off()
```

# Classification

```{r}
set.seed(42)

make_log_feats <- function(df) {
  df %>%
    mutate(
      log_total_spent = log1p(total_spent),
      log_total_qty = log1p(total_qty),
      log_num_orders = log1p(num_orders),
      log_avg_order = log1p(avg_order)
    ) %>%
    select(CustomerID, starts_with("log_"))
}

# split customers first
train_index <- createDataPartition(cust_level_aggregation$CustomerID, p = 0.8, list = FALSE)
cust_train <- cust_level_aggregation[train_index, ]
cust_test <- cust_level_aggregation[-train_index, ]

# log + scale using training stats only
train_feats <- make_log_feats(cust_train)
test_feats <- make_log_feats(cust_test)

train_scaled <- scale(train_feats[, -1])
train_center <- attr(train_scaled, "scaled:center")
train_scale <- attr(train_scaled, "scaled:scale")

test_scaled <- scale(
  test_feats[, -1],
  center = train_center,
  scale = train_scale
)

# k-means fit on training customers only
set.seed(42)
km_train <- kmeans(train_scaled, centers = 3, nstart = 25)

segment_names <- c(
  "High-Value Frequent Buyers",
  "Low-Value Casual Buyers",
  "Large-Basket Buyers"
)

train_segments <- segment_names[km_train$cluster]

assign_to_centroid <- function(mat, centers) {
  apply(mat, 1, function(r) {
    dists <- apply(centers, 1, function(cntr) sum((r - cntr) ^ 2))
    which.min(dists)
  })
}

test_segments <- segment_names[assign_to_centroid(test_scaled, km_train$centers)]

train_data <- cust_train %>%
  mutate(segment = train_segments) %>%
  mutate(label = factor(
    ifelse(segment == "High-Value Frequent Buyers", "High", "Not_High"),
    levels = c("Not_High", "High")
  ))

test_data <- cust_test %>%
  mutate(segment = test_segments) %>%
  mutate(label = factor(
    ifelse(segment == "High-Value Frequent Buyers", "High", "Not_High"),
    levels = c("Not_High", "High")
  ))

# build features
train_x <- train_data %>% select(total_spent, total_qty, num_orders, avg_order)
train_y <- train_data$label

test_x <- test_data %>% select(total_spent, total_qty, num_orders, avg_order)
test_y <- test_data$label

# 1. Decision Tree Classification
model_dt <- train(
  train_x, train_y,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 5)
)

model_dt

pred_dt <- predict(model_dt, test_x)

cm_dt <- confusionMatrix(pred_dt, test_y)
cm_dt

imp_dt <- varImp(model_dt)$importance
imp_dt$Feature <- rownames(imp_dt)

p3 <- ggplot(imp_dt, aes(x = reorder(Feature, Overall), y = Overall, fill = Feature)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Feature Importance for Predicting High-Value Customers",
    subtitle = "Higher score indicates stronger influence in classification decision",
    x = "Customer Features",
    y = "Importance Score"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  ) +
  theme(text = element_text(size = 25))

p3

png("plots/classification_feature_importance.png", width = 1400, height = 1000)
print(p3)
dev.off()
```

# ROC Curve (Decision Tree vs Random Forest)

```{r}
# probability predictions for ROC
dt_prob <- predict(model_dt, test_x, type = "prob")[, "High"]

# 2. Random Forest Classification
# train random forest
set.seed(42)
model_rf <- randomForest(
  label ~ total_spent + total_qty + num_orders + avg_order,
  data = train_data,
  ntree = 300,
  importance = TRUE
)

# RF probabilities
rf_prob <- predict(model_rf, test_x, type = "prob")[, "High"]

# ROC objects
roc_dt <- roc(test_y, dt_prob, levels = c("Not_High", "High"), direction = "<")
roc_rf <- roc(test_y, rf_prob, levels = c("Not_High", "High"), direction = "<")

# plot ROC curves
png("plots/classification_roc_curve.png", width = 1400, height = 1000)

plot(roc_dt, col = "#2E86C1", lwd = 3,
     main = "ROC Curve: Decision Tree vs Random Forest",
     legacy.axes = TRUE)
lines(roc_rf, col = "#28B463", lwd = 3)

auc_dt <- auc(roc_dt)
auc_rf <- auc(roc_rf)

text(0.65, 0.25, paste0("Decision Tree AUC = ", round(auc_dt, 3)),
     col = "#2E86C1", cex = 1.4)
text(0.65, 0.15, paste0("Random Forest AUC = ", round(auc_rf, 3)),
     col = "#28B463", cex = 1.4)

better_model <- ifelse(auc_rf > auc_dt,
                       "Random Forest performs better",
                       "Decision Tree performs better")

text(0.65, 0.05, better_model, col = "black", font = 2, cex = 1.4)

legend("bottomright",
       legend = c("Decision Tree", "Random Forest"),
       col = c("#2E86C1", "#28B463"), lwd = 3)

dev.off()
```

# Random Forest Feature Importance

```{r}
rf_imp <- importance(model_rf)
rf_imp_df <- data.frame(
  Feature = rownames(rf_imp),
  Importance = rf_imp[, 1]
)

p4 <- ggplot(rf_imp_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Random Forest Feature Importance",
    subtitle = "Higher bars mean a stronger contribution to predicting High-Value customers",
    x = "Customer Features",
    y = "Importance Score"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  ) +
  theme(text = element_text(size = 25))

p4

png("plots/classification_rf_importance.png", width = 1400, height = 1000)
print(p4)
dev.off()
```

# Classification Performance Comparison

```{r}
# dt confusion matrix
cm_dt <- confusionMatrix(pred_dt, test_y)

# rf predictions + confusion matrix
pred_rf <- predict(model_rf, test_x)
cm_rf <- confusionMatrix(pred_rf, test_y)

cm_dt
cm_rf
```

```{r}
performance_table <- data.frame(
  Model = c("Decision Tree", "Random Forest"),
  Accuracy = c(cm_dt$overall["Accuracy"], cm_rf$overall["Accuracy"]),
  Sensitivity = c(cm_dt$byClass["Sensitivity"], cm_rf$byClass["Sensitivity"]),
  Specificity = c(cm_dt$byClass["Specificity"], cm_rf$byClass["Specificity"]),
  AUC = c(as.numeric(auc(roc_dt)), as.numeric(auc(roc_rf)))
)

knitr::kable(performance_table, digits = 3,
             caption = "Classification Performance: Decision Tree vs Random Forest")
```

# Final Conclusion

1. Association rules revealed meaningful product pairings based on lift.
2. Clustering created three clear customer segments:
   • High-Value Frequent Buyers  
   • Low-Value Casual Buyers  
   • Large-Basket Buyers  

3. Classification identified key drivers of high-value customers:
   • Decision Tree: total_spent, num_orders strongest predictors  
   • Random Forest achieved higher AUC and accuracy

4. ROC analysis shows Random Forest outperforms Decision Tree consistently.

Overall, the data mining tasks—association mining, clustering, and classification—
provide a complete view of customer behavior and help identify high-value targets.
